---
title: "Lab 1"
number-sections: true
format: 
  html: default
  pdf: default
---

## Overview

::: callout-note
To complete this lab, you will need:

-   access to the shared posit.cloud space for this course (link in Brightspace)
-   a quarto.pub account (free)
-   GitHub Copilot account approved and enabled in posit.cloud
:::

Go to the shared posit.cloud workspace for this class and open the `Labs` project. Open the `lab01.qmd` file and complete the exercises. Below is an annotated guide to assist you. There is also a video in the Brightspace Todo section for this module.

We will be using the `gapminder` data for this lab so let's start by loading the tidyverse family of packages and gapminder. Since this is more of a report, we'll also be using the `message: false` option to suppress the output message from loading `tidyverse`.

```{r}
#| message: false
#| echo: fenced
library(tidyverse)
library(gapminder)
```

## Exercises

There are seven exercises in this lab. Grading is shown in @sec-grading at the end of the document.

### Exercise 1

::: exercise
Using `ggplot` plot the changes in life expectancy over time for the US, Japan, and Russia. Use a line plot. The geom is `geom_line()`
:::

First, let's `glimpse()` the data to remind us what it looks like.

```{r}
glimpse(gapminder)
```

Next, let's check the correct spelling of each country to see how we should reference them, remember that country names may have changed over time (e.g., USSR -\> Russia and other countries). We can use the `pull()` function so we don't get pages of output for a single column and can easily see the names.

```{r}
gapminder |> 
  distinct(country) |> 
  pull()
```

It looks like Russia isn't included in the dataset. Instead of Russia, we'll examine Bulgaria, Japan, United States. We'll create a new tibble called `three_countries` to store our restricted data. Since we are looking at life expectancy changes over time, the three columns we'll need are `country`, `year`, and `lifeExp`.

```{r}
three_countries <-  gapminder |> 
  select(country, year, lifeExp) |> 
  filter(country %in% c("Bulgaria", "Japan", "United States"))
```

Now we can use `ggplot` to create our chart.

```{r}
three_countries |> 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country))
```

### Exercise 2

::: exercise
Modify the code below to add a title to the plot, create a more meaningful y-axis label, and use `theme_minimal()` to change the theme of the plot.
:::

Refer to the [Getting Started](https://jsuleiman.com/) chapter, which is when we first labeled the axes and added a title to plots.

```{r}
three_countries |> 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country))
```

### Exercise 3

::: exercise
In the empty code chunk below, create the same graphic using three new countries of your choosing. Set the chunk option to render the plot, but not the code.
:::

```{r}
# insert code here
```

### Exercise 4

::: exercise
Try to recreate the plot **and the linked reference to it** shown below.

```{r}
#| label: fig-three_life_exp
#| fig-cap: "Life expectancy trend - Bulgaria, Japan, United States"
#| echo: false
three_countries |> 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country)) +
  labs(
    title = "Life Expectancy Over Time",
    x = "Year",
    y = "Life Expectancy",
    color = "Country"
  ) +
  theme_minimal()
```

You can see in @fig-three_life_exp that Bulgaria's life expectancy increase slowed starting in the mid 1960s.
:::

::: callout-tip
## Hints

-   use the chunk option `fig-cap: "Write your caption here"` to replicate the caption towards the bottom of the plot.

-   use the chunk option `label: fig-unique_label_name` to make a label that becomes cross-referenceable. Ideally, `unique_label_name` should be something unique and meaningful like `three_country_life_exp`. The `fig-` prefix is required to properly cross reference figures.

-   After the code chunk with the label has been run or rendered, in the visual editor, you should be able to select Insert -\> Cross Reference to place it in the narrative text. It will default as link to "Figure 1" which is the first labeled figure in the document.
:::

```{r}
# insert code here
```

### Exercise 5

::: exercise
Copilot doesn't suggest completions when in Visual mode.

Insert a code chunk below using **Insert -\> Exectuable Cell -\> R** in the visual editor, then switch to Source mode.

While in source mode, type the following comment line: `# Create a plot of life expectancy over time for the United States and Portugal`.

In the narrative below the plot, describe how well Copilot did.
:::

### Exercise 6

::: exercise
In the YAML section at the top of the document, replace the author's name with your own.
:::

### Exercise 7

::: exercise
Publish your document to quarto.pub and copy/paste the link to the Lab 1 assignment in Brightspace.
:::

## Grading {#sec-grading}

| Exercise   | Points |
|------------|--------|
| Exercise 1 | 10     |
| Exercise 2 | 10     |
| Exercise 3 | 10     |
| Exercise 4 | 10     |
| Exercise 5 | 10     |
| Exercise 6 | 10     |
| Exercise 7 | 40     |
| **Total**  | 100    |
